EBASE manuscript

### 2022-12-20 manu-draft.qmd
working on ebase theory

@@

@@

The standard prior distributions for the EBASE parameters follow normal distributions with defined means and standard deviations.  The sensitivity analysis evaluated changes in the distributions by varying the standard deviations from small to large to test the effect of more or less constrained ranges, respectively, on the priors. For each parameter, low, medium, and high values for the standard deviation were assessed as 10%, 100%, and 1000% of the default value.  Each unique combination of low, medium, and high values for the standard deviation of each parameter were evaluated, creating 3$^3$ = 27 different combinations of prior distributions that were used with the simulated time series in EBASE.  Because initial assessments indicated that all parameter estimates were especially sensitive to changes in the prior distribution for the $b$ parameter, a second sensitivity analysis evaluated additional changes in both the mean and standard deviation of the prior for the $b$ parameter.  Changes in both the mean and standard deviation of the $b$ prior were evaluated, where the mean was varied as 50%, 100%, and 200% (the upper limit) of the default and the standard deviation was varied as before.  This resulted in 3$^2$ = 9 unique combinations of means and standard deviations for the $b$ prior that were evaluated.

The optimization period used for a time series determines how much data are used to estimate the three EBASE parameters ($a$, $r$, and $b$).  Unlike BASEmetab, the EBASE software allows flexibility in the period of time used for optimization, with a minimum period of time being one day and a maximum period of being the length of the time series.  For a chosen optimization period, a single estimate for $r$ and $b$ is returned by the model, whereas $n$ number of observations in each optimization period are returned for the $a$ parameter (e.g., 96 values for  24 hour time period if the samplign interval is 15 minutes).  Short periods of observation may return unstable parameter estimates if forcing factors (e.g., wind, temperature) that affect estuarine biogeochemical rates occur at time scales longer the chosen period of optimization for EBASE.  As such, potentially longer optimization periods may return more stable and robust estimates given that more data are used for parameter estimation.  An optimization period of one day was compared to results using an optimization period of seven days for the simulated one-year time series. Each optimization period was evaluated for every unique combination of prior distributions, totaling 54 unique analyses for the assessment of changing standard deviations of all parameters and 18 unique analyses for the assessment of changing the mean and standard deviation of only the $b$ parameter.

The metabolic estimates returned from each analysis were compared to the known values to determine which priors the model was most sensitive to and if the conclusions varied by optimization period.  Estimates for the modeled dissolved oxygen, gross production, respiration ($r$), gas exchange ($D$), and $a$ were compared to those from the simulated time series using coefficient of variation ($R^2$) and root mean square error ($RMSE$) summaries of the 1:1 comparison of the estimated vs simulated values.  For the one day optimization period, all parameter estimates returned at the time step of the input data (e.g., $a$) were daily-averaged for comparison.  For the seven day optimization period, all parameter estimates and those from the known time series were weekly-averaged.  Estimates for the $b$ parameter were evaluated only by visual comparison because a single value was used in the simulated time series, preventing the estimate of quantitative comparisons.

>>>>

The Estuarine BAyesian Single station Estimation (EBASE) method provides a simple approach to quantify metabolic parameters from continuous dissolved oxygen time series.  The estimates are based on a mass balance equation described in @Grace15 that is modified using a gas exchange calculation described in @Wanninkhof14 (@eq-ebase).

$$
\frac{\delta{C_d}}{\delta{t}} = [\,aPAR]\, - [\,r]\, - \frac{1}{H}\left[-bU_{10}^2\left(\frac{s_c}{600} \right)^{-0.5} \left(C_{Sat} - C_d \right )\right]
$$ {#eq-ebase}

The metabolic estimates are defined by the change in dissolved oxygen over the time step $\frac{\delta{C_d}}{\delta{t}}$, where gross production is provided by $aPAR$, respiration is provided by $r$, and gas exchange is provided by the remainder [as in @Wanninkhof14].  Required data for the equation are dissolved oxygen concentration as $C_d$ (mmol/m$^3$), solar radiation as $PAR$ (W/m$^2$/s), water column depth as $H$ (m), and wind speed at 10 m height as $U$ (m/s).  The schmidt number $s_c$  is the ratio of the kinematic viscosity of water to the molecular diffusivity of oxygen calculated as a function of water tempererature and salinity using the polynomial wit of @Wanninkhof14.

Dissolved oxygen at saturation $C_{Sat}$ is calculated from the input data using @Garcia92 as a function of water temperature (C, $T$) and salinity (psu, $sal$). @eq-ts is temperature scaled, as input to @eq-co that calculates the solubility of dissolved oxygen per volume of seawater at the given temperature. The coefficients are A$_0$ = 5.80818, A$_1$ = 3.20684, A$_2$ = 4.11890, A$_3$ = 4.93845, A$_4$ = 1.01567, A$_5$ = 1.41575, B$_0$ = -7.01211e-3, B$_1$ = -7.25958e-3, B$_2$ = -7.93334e-3, B$_3$ = -5.54491e-3, and C$_0$ = -1.32412e-7.

$$
T_s = \log((298.15 - T)/(273.15 + T)
$$ {#eq-ts}
$$
C_O = A_0 + A_1 \cdot T_s + A_2 \cdot T_s^2 + A_3 \cdot T_s^3 + A_4 \cdot T_s^4 + A_5 \cdot T_s^5 + sal \cdot (B_0 + B_1 \cdot T_s + B_2 \cdot T_s^2 + B_3 \cdot T_s^3) + C_0 \cdot sal^2
$$ {#eq-co}

Dissolved oxygen saturation is then calculated by multiplying $C_o$ by the density of seawater given temperature and salinity (@eq-csat). The parameter $\rho$ (kg/m$^3$) is seawater density at zero elevation given temperature and salinity [@Millero81].

$$
C_{sat} = \exp(C_O) \cdot \rho_{T, sal} \cdot 1e-3
$$ {#eq-csat}

The remaining three parameters $a$, $r$, and $b$ are estimated by likelihood given the observed data with the JAGS model using prior distributions shown in the model file. At each time step, the change in oxygen concentration between time steps is calculated from the equation using model inputs and parameter guesses, and then a finite difference approximation is used to estimate modeled oxygen concentration.  The estimated concentration is also returned, which can be compared to observed as one measure of model performance.

@@

The standard prior distributions for the EBASE parameters follow normal distributions with defined means and standard deviations.  The sensitivity analysis evaluated changes in the distributions by varying the standard deviations from small to large to test the effect of more or less constrained ranges, respectively, on the priors. For each parameter, low, medium, and high values for the standard deviation were assessed as 10%, 100%, and 1000% of the default values.  Each unique combination of low, medium, and high values for the standard deviation of each parameter were evaluated, creating 3$^3$ = 27 different combinations of prior distributions that were used with the simulated time series in EBASE.  Because initial assessments indicated that all parameter estimates were especially sensitive to changes in the prior distribution for the $b$ parameter, a second sensitivity analysis evaluated additional changes in both the mean and standard deviation of the prior.  The mean of the $b$ prior was varied as 50%, 100%, and 200% (the upper limit used in the software) of the default and the standard deviation was varied as before.  This resulted in 3$^2$ = 9 unique combinations of means and standard deviations for the $b$ prior that were evaluated.

The optimization period used for a time series determines how much data are used to estimate the three EBASE parameters ($a$, $r$, and $b$).  Unlike BASEmetab, the EBASE software allows flexibility in the period of time used for optimization, with a minimum period of time being one day and a maximum period of being the length of the time series.  For a chosen optimization period, a single estimate for $r$ and $b$ is returned by the model, whereas $n$ number of observations equal to the number of time steps in each optimization period are returned for the $a$ parameter (e.g., 96 values for  24 hour time period if the sampling interval is 15 minutes).  Short periods of observation may return unstable parameter estimates if forcing factors (e.g., wind, temperature) that affect estuarine biogeochemical rates occur at time scales longer the chosen period of optimization for EBASE.  As such, potentially longer optimization periods may return more stable and robust estimates given that more data are used for parameter estimation.  An optimization period of one day was compared to results using an optimization period of seven days for the simulated one-year time series. Each optimization period was evaluated for every unique combination of prior distributions, totaling 54 unique analyses for the assessment of changing standard deviations of all parameters and 18 unique analyses for the assessment of changing the mean and standard deviation of only the $b$ parameter.

The metabolic estimates returned from each analysis were compared to the known values to identify priors the model was most sensitive to and if the conclusions varied by optimization period.  Estimates for the modeled dissolved oxygen, gross production, respiration ($r$), gas exchange ($D$), and $a$ were compared to those from the simulated time series using coefficient of variation ($R^2$) and root mean square error ($RMSE$) summaries of the 1:1 comparison of the estimated vs simulated values.  For the one day optimization period, all parameter estimates returned at the time step of the input data (i.e., modeled dissoled oxygen, $D$, and $a$) were daily-averaged for comparison.  For the seven day optimization period, all parameter estimates and those from the known time series were weekly-averaged.  Estimates for the $b$ parameter were evaluated only by visual comparison because a single value was used in the simulated time series, preventing the estimate of quantitative comparisons.

@@
